---
title: "Correlation"
author: "morgan moyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(lme4)
library(lmerTest)
library(multcomp) # not available for this version of R
library(stringr)
library(textstem)
library(tidyverse)
library(kableExtra)
library(emmeans)
library(BayesFactor)
library(ggrepel)
theme_set(theme_bw())
cbPalette <- c("#56B4E9", "#D55E00", "#009E73","#999999", "#E69F00")
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
source("../../helpers.R")

sc <- read_csv("../data/categorization_agg.csv")  
names(sc)[names(sc) == 'Experiment'] <- 'ConceptualFoil'
sj <- read_csv("../../01_experiment/total/data/similarity_agg.csv")
names(sj)[names(sj) == 'Experiment'] <- 'ConceptualFoil'
names(sj)[names(sj) == 'Response'] <- 'SimilarityRating'
```


```{r}
names(sc)
# Center first
sc.cen <- sc %>% 
  mutate(Task = ifelse(Task == "Valence", "Valence", "Conceptual"),
          Task = factor(Task, levels = c("Valence", "Conceptual"))) %>% 
  mutate(Task = as.factor(Task),
         ConceptualFoil = as.factor(ConceptualFoil),
         cTask = as.numeric(Task)-mean(as.numeric(Task)))

# Fit model (example with lm, adjust if you're using lmer)
mod1 <- lmer(MeanReactionTime ~ cTask * ConceptualFoil + (1|), data = sc.cen)

# Task contrast within each foil
task_by_foil <- emmeans(mod1, pairwise ~ cTask | ConceptualFoil)
task_by_foil$contrasts
task_effects <- as.data.frame(task_by_foil$contrasts)
emms1 <- emmeans(mod1, ~ ConceptualFoil)
```


```{r}
sj_reduced <- sj %>%
  filter(FeatureMatch %in% c("ConceptualMatchingOnly","ValenceMatchingOnly")) %>% 
  mutate(FeatureMatch = as.factor(FeatureMatch),
         ConceptualFoil = as.factor(ConceptualFoil),
         cFeatureMatch = as.numeric(FeatureMatch)-mean(as.numeric(FeatureMatch)))# keep only relevant cases

```


```{r}

mod2 <- lmer(SimilarityRating ~ cFeatureMatch * ConceptualFoil + (1|ID.true), data = sj_reduced)

# Example: pairwise contrasts for FeatureMatch within each foil
fm_by_foil <- emmeans(mod2, pairwise ~ cFeatureMatch | ConceptualFoil)
fm_effects <- as.data.frame(fm_by_foil$contrasts)
emms2 <- emmeans(mod2, ~ ConceptualFoil)

```

# General Correlation

```{r}

merged <- task_effects %>%
  select(ConceptualFoil, task_est = estimate) %>%
  inner_join(fm_effects %>% select(ConceptualFoil, fm_est = estimate),
             by = "ConceptualFoil")

```

Across the seven Conceptual Foil conditions, the size of the Task effect on log-transformed reaction times in Study 1 was strongly correlated with the size of the Match–NonMatch effect on similarity judgments in Study 2, r(5) = 0.96, 95% CI [0.66, 0.996], p = .0026.
```{r}
cor.test(merged$task_est, merged$fm_est, method = "pearson")

```


# Directional Prediction:
do lower RTs in Valence trials predict higher similarity judgments in ValenceMatchingOnly trials across foils

IV (Study 1): Foil-wise mean log RT for Task == Valence.

Lower = faster responses.

DV (Study 2): Foil-wise mean similarity rating for FeatureMatch == ValenceMatchingOnly.

Higher = stronger similarity judgment.

Prediction: Faster RTs (smaller log RTs) → Higher similarity ratings.

This implies a negative correlation (because lower RTs = higher ratings).

```{r}

# Study 1
# Get Valence-only EMMs per foil
emms_valence <- emmeans(mod1, ~ ConceptualFoil, at = list(Task = "Valence"))
df_valence <- as.data.frame(emms_valence) %>%
  select(ConceptualFoil, valence_rt = emmean)

# Study 2
# Get ValenceMatchingOnly EMMs per foil
emms_valmatch <- emmeans(mod2, ~ ConceptualFoil, at = list(FeatureMatch = "ValenceMatchingOnly"))
df_valmatch <- as.data.frame(emms_valmatch) %>%
  select(ConceptualFoil, valmatch_sim = emmean)


# Merge
merged <- inner_join(df_valence, df_valmatch, by = "ConceptualFoil")


```

## Use a one-sided correlation test

```{r}
cor.test(merged$valence_rt, merged$valmatch_sim,
         alternative = "less",   # tests negative correlation
         method = "pearson")

```

A one-tailed correlation test between foil-wise Valence RTs (Study 1) and ValenceMatchingOnly similarity ratings (Study 2) yielded a moderate negative correlation (r = –.49, 95% CI [–1.00, .40]), consistent with the directional hypothesis, but not statistically significant, t(4) = –1.11, p = .16.

## Visualize
```{r}
ggplot(merged, aes(x = valence_rt, y = valmatch_sim, label = ConceptualFoil)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text(vjust = -1, size = 3) +
  labs(
    x = "Mean log RT (Task = Valence)",
    y = "Mean similarity (FeatureMatch = ValenceMatchingOnly)",
    title = "Relationship between Valence RTs and ValenceMatchingOnly similarity"
  ) +
  theme_minimal(base_size = 14)
```


## Bayes factor
H0: correlation = 0

H1: correlation < 0 (restricted interval)

If BF > 3 → moderate evidence for H1,
If BF < 1/3 → moderate evidence for H0,
If between → inconclusive.
```{r}
bf <- correlationBF(merged$valence_rt, merged$valmatch_sim,
                    nullInterval = c(-1, 0))  # directional: negative correlation
bf

```

Bayes factor analysis
--------------
[1] Alt., r=0.333 -1<rho<0    : 1.420574 ±NA%
[2] Alt., r=0.333 !(-1<rho<0) : 0.472915 ±NA%

Against denominator:
  Null, rho = 0 
---
Bayes factor type: BFcorrelation, Jeffreys-beta*



